Likelihoods for a general class of ARGs under the SMC

### 2025-04-29 paper.tex
Fixes

In Fig.~\ref{fig:lik-surface} we begin with 
with a sample from the SMC model 
containing complete information about every simulated event
(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). \revpoint{2}{30}
We then plot the likelihood computed by Eq.~\eqref{eq:full-lik}
as we vary the recombination rate parameter $r$, as a ratio
of the likelihood at the true value of $r$.
Reassuringly, we can
see that the minimum of the likelihood-ratio curve when using the
full simulated ARG is close to the true value of the parameter.
(The likelihood ratio curve obtained when computing the likelihood
under the full coalescent with recombination model using the KYF
formula is identical; not shown.)
We next compute the likelihood ratio for the ARG obtained when we
simplify~\citep{kelleher_efficient_2018,wong_general_2023}
 out the recombination nodes from the full ARG
(``RE nodes removed''; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime
uses distinct nodes for the left and right parent of each recombination).
The curve we obtain is indistinguishable from that of the full
ARG, demonstrating that the locally unary portions of coalescent nodes
contain all the information that we need about recombination events
for this application. \revpoint{2}{31}
When we fully simplify the simulated ARG
to remove all locally unary portions of nodes from trees
(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees)  
we can see there is a significant loss of information, \revpoint{2}{32}
and the minimum of the
likelihood ratio surface is biased away from its true value,
as expected, because removing the unary portions of nodes has removed some of the area
of the ARG in which recombination did not occur.
(See~\cite{wong_general_2023} for more discussion on varying
degrees of ARG simplification.)

Fig.~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
robust to the presences of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent
is not unary. Thus, while deleting 25\% of the internal nodes
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary
nodes per tree).
%     name                  leaves  unary       binary  poly
% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
% 1   unary_coal            200.0   456.396522  199.000000  0.000000
% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the
effects on the likelihood across the three examples are not entirely comparable.
For the ARGs containing unary nodes,
removing 25\% of the internal
nodes has negligible effect on the likelihood curve in Fig.~\ref{fig:lik-surface},
and even on the fully simplified
ARG has a very minor effect on location of the minimum.
Supplemental Fig.~S1 explores more extreme examples in which we remove
25\%, 50\% and 75\% of the internal nodes in these ARGs, and demonstrates 
a remarkable robustness to the loss of the majority of the internal
nodes in these examples. Interestingly, the loss of 75\% of the 
internal nodes  in the fully simplified ARG 
(1841 of 2455, leading  to an average of 27 polytomies per tree)
appears to improve performance, substantially 
reducing the bias away from the true parameter value.

@@

\begin{figure}[!ht]
\centering
\includegraphics{figures/likelihood_surface_suppl}
    \caption{Log likelihood-ratio curves for three different ARGs 
    with increasing fractions of polytomies.
    The base ARGs are as described in Fig.~3, but here we remove 
    25\%, 50\% and 75\% of the internal nodes to create polytomies.
    The true value of recombination parameter is indicated by the 
    dashed line, as in Fig.~3.
    \label{fig:likelihood_suppl}}
\end{figure}

>>>>

In Fig.~\ref{fig:lik-surface} we begin 
with a sample from the SMC model 
containing complete information about every simulated event
(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). \revpoint{2}{30}
We then plot the likelihood computed by Eq.~\eqref{eq:full-lik}
as we vary the recombination rate parameter $r$, as a ratio
of the likelihood at the true value of $r$.
Reassuringly, we can
see that the minimum of the likelihood-ratio curve when using the
full simulated ARG is close to the true value of the parameter.
(The likelihood ratio curve obtained when computing the likelihood
under the full coalescent with recombination model using the KYF
formula is identical; not shown.)
We next compute the likelihood ratio for the ARG obtained when we
simplify~\citep{kelleher_efficient_2018,wong_general_2023}
 out the recombination nodes from the full ARG
(``RE nodes removed''; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime
uses distinct nodes for the left and right parent of each recombination).
The curve we obtain is indistinguishable from that of the full
ARG, demonstrating that the locally unary portions of coalescent nodes
contain all the information that we need about recombination events
for this application. \revpoint{2}{31}
When we fully simplify the simulated ARG
to remove all locally unary portions of nodes from trees
(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees)  
we can see there is a significant loss of information, \revpoint{2}{32}
and the minimum of the
likelihood ratio surface is biased away from its true value,
as expected, because removing the unary portions of nodes has removed some of the area
of the ARG in which recombination did not occur.
(See~\cite{wong_general_2023} for more discussion on varying
degrees of ARG simplification.)

Fig.~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
robust to the presences of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent
is not unary. Thus, while deleting 25\% of the internal nodes
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary
nodes per tree).
%     name                  leaves  unary       binary  poly
% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
% 1   unary_coal            200.0   456.396522  199.000000  0.000000
% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the
effects on the likelihood across the three examples are not entirely comparable.
For the ARGs containing unary nodes,
removing 25\% of the internal
nodes has negligible effect on the likelihood curve in Fig.~\ref{fig:lik-surface},
and even on the fully simplified
ARG has a very minor effect on the location of the minimum.
Supplemental Fig.~S1 explores more extreme examples in which we remove
25\%, 50\% and 75\% of the internal nodes in these ARGs, and demonstrates 
a remarkable robustness to the loss of the majority of the internal
nodes in these examples. Interestingly, the loss of 75\% of the 
internal nodes  in the fully simplified ARG 
(1841 of 2455, leading  to an average of 27 polytomies per tree)
appears to improve performance, substantially 
reducing the bias away from the true parameter value.

@@

\begin{figure}[!ht]
\centering
\includegraphics{figures/likelihood_surface_suppl}
    \caption{Log likelihood-ratio curves for three different ARGs 
    with increasing fractions of polytomies.
    The base ARGs are as described in Fig.~3, but here we remove 
    25\%, 50\% and 75\% of the internal nodes to create polytomies.
    The true value of the recombination parameter is indicated by the 
    dashed line, as in Fig.~3.
    \label{fig:likelihood_suppl}}
\end{figure}
